Formation and evolution of intermediate-mass ratio inspirals.

### 2020-07-13 main.tex
final remarks added

\titlerunning{Formation of IMRIs in Milky Way globular clusters and the Local Volume}	
\authorrunning{Arca Sedda, Amaro-Seoane, Chen}
\author{Manuel Arca Sedda \inst{1} \and Pau Amaro-Seoane \inst{2,}\inst{3,}\inst{4,}\inst{5} \and Xian Chen \inst{6,}\inst{5} }

\institute{Astronomisches Rechen-Institut. Zentrum f\"{u}r Astronomie der Universit\"{a}t Heidelberg, M\"onchhofstrasse 12-14, Heidelberg, D-69120, DE              
         \and         
			  Institute of Space Sciences (ICE, CSIC) \& Institut d'Estudis Espacials de Catalunya (IEEC) at Campus UAB, Carrer de Can Magrans s/n 08193 Barcelona, Spain
		\and         
			  Institute of Applied Mathematics, Academy of Mathematics and Systems Science, CAS, Beijing 100190, China
		\and
			  Zentrum f{\"u}r Astronomie und Astrophysik, TU Berlin, Hardenbergstra{\ss}e 36, 10623 Berlin, Germany
		\and         
			  Kavli Institute for Astronomy and Astrophysics at Peking University, Beijing 100871, P.R. China	
		\and
			Astronomy Department, School of Physics, Peking University, Beijing 100871, P.R. China	             
\\		\email{m.arcasedda@gmail.com}			
\\    	\email{pau@ice.cat}
\\		\email{xian.chen@pku.edu.cn }
}

@@

Intermediate mass black holes (IMBH), with masses in the range $10^2-10^5\Ms$, might represent the missing link between stellar and supermassive mass black holes. Dense stellar systems, such as globular clusters (GCs), are thought to be ideal factories for IMBH production via formation and collapse of a very massive stars through stellar collisions \citep{zwart02, freitag06b,freitag06c, giersz15, mapelli16},
or via multiple interactions and mergers between stars and stellar-mass BHs \citep{giersz15}. Unfortunately, a striking observational evidence for the presence of IMBHs inhabiting GCs is still missing due to the little dynamical effects that these objects have on their surroundings \citep[for recent reviews see][]{mezcua17,greene19}. Indeed, models suggest that several processes can mimic an IMBH, like anisotropies inthe GC kinematic  properties \citep{zocchi}, or the presence of a dense cluster of stellar mass BHs -- namely a BH subsystem -- that dominate the inner cluster centre \citep{AAG18b,AS16,vandermarel10}. Nevertheless, a few observational IMBHs candidates have been found for Galactic GCs

@@

properties \citep{AAG18a}. Due to this, finding an unique way to unravel the presence of IMBHs in GCs represents one of the most interesting challenges in modern astronomy. Beside this, the presence of an IMBH sitting in the centre of a dense cluster represents a scenario particularly appealing from the perspectives of gravitational waves (GW) astronomy. Indeed, a compact remnants orbiting the IMBH can enter a regime where GWs emission dominates, leading to the formation of an intermediate mass-ratio inspiral \citep[IMRI,][]{konstantinidis13,leigh14,haster16,macleod16}, a class of sources possible audible with the next generation of GW observatories like LISA \citep{seoane07,amaro12,seoane18}. However, in the highly dense regions that characterize star clusters centres, the formation of IMRIs is not a smooth process. Indeed, due to the continuous interactions with stars, an IMRI ``progenitor'', namely a tight IMBH-BH binary, might be subjected to strong perturbation induced, for instance by a passing-by BH. The three-body interaction involving the IMBH and the two BHs can lead to a variety of end states, including the formation of an IMRI, a stellar BH binary, the ejection of one BH, or even both, or the development of a head-on collision. At some extent, this scenario is similar to what is expected to happen in galactic nuclei, where supermassive BHs (SMBHs) can capture a compact object to form an extreme mass ratio inspiral. However, in the case of IMRIs the picture is complicated by the fact that this chaotic process can transfer to the IMBH an amount of energy sufficient to displace it sensibly from the cluster centre. Differently from galactic nuclei, where the SMBH remains well seated in the galactic potential well, the IMBH motion makes hard the use of any analytical approach to solve IMRIs dynamics. Quantifying the branching ratio for IMRIs formation constitute a fundamental step to assess the probability to observe these GW sources with the next generation of space-based detectors like LISA\footnote{\url{https://www.elisascience.org/}} \citep{seoane07,amaro12}, TianQin \citep{tianqin16} or Taiji \citep{taiji17}. In this paper, we model the potential formation of an IMRI from the interactions between an IMBH and two stellar mass BHs. This type of  configuration can temporarily form in the centre of clusters with IMBH mass $\sim 100-1000\Ms$ \citep{konstantinidis13,macleod16} and could last up to $10^7$ yr \citep{macleod16} although this depends strongly on the cluster initial properties. In fact, it is difficult to predict the actual amount of BHs that, at any time, interact with the IMBH, primarily because the majority of the studies conducted in the literature assume that the IMBH is already at the centre of the host cluster when stellar BHs form and sink toward the cluster centre \citep{leigh14,haster16}. Models in which the IMBH is a byproduct of stellar evolution and collisions, indeed, seem to point out that after the IMBH grows only a handful of stellar BHs are still present in the cluster $<3-5$ \citep{AAG19}. To reach the aim, we use $N$-body simulations that take into account in particles' equations of motion both the star cluster gravitational potential and post-Newtonian corrections at 1, 2, and 2.5 order. Varying the IMBH and BHs masses, their orbital configuration, and the host cluster structural properties, we build-up three sets consisting of 2000 models each, which allow us to uncover three possible scenarios for IMRIs formation.  

@@

{\bf
We simulate the evolution of the innermost region of the cluster, modelling the IMBH and two stellar mass BHs as live particles and
the remaining cluster as an external static potential. In order to explore the parameter space, we create six different models, each one consisting 
of 4000 simulations gathered in four different sub-classes, depending on the IMBH mass, for a total of 48,000 simulations. Our model is sketched in Figure \ref{f1}.

\begin{figure}
    \centering
    \includegraphics[width=7cm]{triple}
    \caption{Sketch of the IMBH-BH-BH triple configuration.}    
\label{f1}
\end{figure}

We vary the central IMBH mass in four classes, namely ${\rm Log}(M_\ibh/\Ms) = 2-3-4-5$. This range covers typical values of putative IMBH masses forming in stellar systems of various sizes, from young and open clusters, to globular clusters, and up to low-mass nuclear clusters and dwarf galaxies.
We note that the lowest value taken for $M_\ibh$ will likely involve mergers with mass ratio $> 0.1$, thus they fall outside the range of IMRIs. Nonetheless, exploring the low end of IMBH mass function will help us in better understanding the perspectives of IMBH-BH mergers from the point of view of both low- and high-frequency GW detectors, as discussed in Section \ref{gra}.

@@

These results have two main implications. On the one hand, the IMRI secondary mass tells us something about the mass spectrum of BHs lurking around an IMBH. On the other hand, it provides clues about the IMBH formation process. In fact, if the IMBH formed during cluster's early life, we might expect that the mass spectrum of IMRIs secondary maps roughly the stellar BH mass spectrum. If, instead, dynamical interactions have enough time to shape the BHs reservoir, we might expect that the mass spectrum of IMRI secondaries will reflect the outcome of stellar dynamics.

\begin{figure*}
\centering
\includegraphics[width=0.8\textwidth]{massdis}\\
%\includegraphics[width=0.8\textwidth]{massdisD}\\
    \caption{%Top panel: 
    Mass distribution of the IMRI secondary for SET0 to 5. Each color corresponds to a different set to facilitate the comparison between different panels and figures. 
    %Bottom panel: the same as above, but differentiating for different IMBH masses: $10^2\Ms$ (straight line), $10^3\Ms$ (dashed line), $10^4\Ms$ (dashed dotted line), $10^5\Ms$ (dotted line) as indicated in the legend. Each color corresponds to a different set to facilitate the comparison between the upper and lower panels.
    }
	\label{fig:f8}
\end{figure*}

@@

\subsection{IMBH survival in dense star clusters}
{\bf
Promptly after a merger event, the anysotropic emission of GWs can impart a recoil kick to the merger product, depending on the mass and spin of the two merging objects \citep{campanelli07,gonzalez07,lousto08,lousto12}. Depending on the kick amplitude, the recoil can kick out the IMBH from the cluster \citep[e.g.][]{bockelmann08}, leading it to wander forever outside the cluster. 
This can significantly affect the retention probability of low-velocity dispersion ($5$ km/s) star clusters, which is limited to $1-5\%$ for BHs with mass $m_\bh \sim 100 \Ms$ that undergo 1-2 consecutive mergers \citep{arca20}.

Assessing the retention probability for IMBHs is crucial to place constraints on their possible presence in star clusters. To explore this aspect, we use our models to create a catalogue of IMRI mergers proceeding on a step-by-step basis as follows: 
\begin{enumerate}
\item we divide the IMBH mass range -- $M_\ibh = 10^2-5\times 10^5\Ms$ -- in 15 values evenly distributed in logarithmic values;
\item for each IMBH mass, we create a sample of 100 stellar BHs with masses evenly drawn in the $M_\bh = 3.5-80 \Ms$ range;
\item we associate to the IMBH and BH a spin randomly drawn between 0-1;
\item we use \cite{jimenez17} numerical relativity fitting formulae to calculate the IMRI merger remnant mass $M_r$, spin $S_r$, and recoil velocity $v_r$, following the procedure depicted in \cite{arca20}.
\end{enumerate}
For each $M_\ibh -m_\bh$ pair, we repeat steps 3 and 4 100 times and, for each of them, we compare the recoil velocity with the escape velocity from the cluster centre, which we set to $\sigma = 5$ or $50$ km/s.

@@

Figure \ref{fig:f11a} shows the value of the remnant IMBH spin as a function of the IMRI mass ratio.

We see that as long as the mass ratio remains below $10^{-2}$ -- which implies an $M_\ibh > 10^4 \Ms$ -- the IMBH spin is not affected by the merger. However, for IMRIs with a mass ratio > 0.01, the remnant IMBH spin can change significantly, with a spin variation $\sim \pm 0.1$ for mass ratios in the range $q = 0.01-0.1$. At mass ratio $q>0.1$, the remnant IMBH spin ($S_r \sim 0.8\pm 0.2$) tends to attain a value close to the average value assumed for the BH spin distribution.
The peculiar dependence between the remnant spin and the IMRI mass ratio has important consequences for the IMBH evolution. One the one hand, if the IMBH formed from the collapse of a gaseous cloud or a very massive star born out of stellar collisions and reach masses $\gtrsim 10^4\Ms$, measuring its spin in a IMRI merger would provide us with information on the processes that regulated the IMBH very formation. On the other hand,
if the IMBH formed via multiple mergers in a ''BH-rich'' stellar environment, the remnant spin can provide clues on the spin distribution of the underlying population of stellar BHs.

To further investigate this point, we explore the effects of a series of subsequent mergers on the IMBH final spin. We assume $M_\ibh = 10^3\Ms$, i.e. a mass such to ensure a retention probability of at least $50\%$ in an IMRI merger event involving a BH smaller than $40\Ms$ (see Figures  \ref{fig:survival} and \ref{fig:f11a}). We set the IMBH spin to either $S_\ibh = 0.1-0.5-0.9$. For each spin value, we model ten consecutive mergers with stellar BHs with masses $3-60\Ms$ and spin drawn from a Gaussian distribution with average $S_\bh = 0.5$ and dispersion 0.1. The resulting merger sequence represents an {\it IMBH merger tree}. For each value of the spin we thus create 100 merger trees and calculate the IMBH spin at each successive merger as shown in Figure \ref{fig:f11b}.
We note that all models tend to converge toward the mean value of the adopted spin distribution for stellar BHs, thus suggesting that it could be possible to link the IMBH spin with the overall spin distribution of stellar BHs inhabiting the star cluster. In fact, complementing a measure of the IMBH spin with the spin distribution inferred from stellar BH mergers observed as GW sources can help in unravelling the IMBH evolution, whether it was driven by mergers with compact objects or by stellar feeding.
}

@@

\subsection{Gravitational wave signal}
{\bf
Our simulations encompass a wide range of IMRI models, in fact touching the layer of ``ordinary'' BH binaries (mass ratio $q>0.1$) and 
scratching the limit of extreme-mass ratio inspirals ($q < 3 \times10^{-5}$). This setup implies that the GW emission connected with our modelled mergers can populate a wide range of frequencies, from milli- to deca-Hz. Figure \ref{fig:f13} show the characteristic amplitude \citep{kocsis12,ASLiK18,seoane18,robson19} as a function of the frequency for a sample of IMRI mergers with different IMBH mass, assuming they are located at redshift $z=0.1$ ($\sim 460$ Mpc) and an observation time of 4 yr. We compare the modelled signal with the sensitivity curve of several detectors, namely the laser interferometer space antenna \citep[LISA,][]{eLISA13,amaro17lisa}, the advanced laser interferometer antenna \citep[ALIA,][]{bender13}, the decihertz observatory \citep[DO,]{arca19}, the Deciherz Gravitational-Wave Observator \citep[DECIGO,]{Kawamura11}, LIGO \citep{abbott16}, and the Einstein Telescope \citep[ET]{punturo10}. We see that mergers involving IMBHs with masses $< 10^5\Ms$ are promising multiband sources that can be seen in one detector during the inspiral phase and in another during the merger. The number of regions in which a detector is sensitive traversed by an IMRI widens at increasing the IMBH mass, in fact in principle an IMBH with mass $<10^3\Ms$ can be seen across the whole milli-decaHz frequency range.

@@

The term $\derd M_\gc / \derd M_\ibh$ can be calculated by inverting Equation \ref{mcmibh} and performing the derivative. For the number distribution of the GCs mass in a given galaxy, $\derd n/\derd M_\gc$, we assume a power-law
\begin{equation}
\frac{\derd n}{\derd M_\gc} = k M_\gc^{-s},
\end{equation}
with the slope $s = 2.2$ and the normalization constant 
\begin{equation*}
k = \frac{\delta M_g (2-s)}{(M_{\gc 2}^{2-s} - M_{\gc 1}^{2-s})}.
\end{equation*}
Assuming Galactic values for the galaxy stellar mass, $M_g = 6\times 10^{10}\Ms$, and star clusters mass range, $M_{\gc 1,2} = (5\times 10^3 - 8\times 10^6) \Ms$, we calculate the corresponding IMBH mass range $M_\ibh \simeq (30 - 4.6\times 10^4)\Ms$ according to Equation \ref{mcmibh}, which can also be used to write $M_\gc^{-s} = aM_\ibh^{-bs}$\footnote{The parameters $a,b$ are obtained manipulating Equation \ref{mcmibh}.}. The $\derd n/(\derd M_g\derd z)$ is obtained exploiting the results in \cite{conselice16}, who studied the distribution of galaxies with stellar masses up to $10^{12}\Ms$ up to redshift $z=8$. In particular, we exploit the following parametric expression of galaxies number density 
\begin{equation}
\phi(z) = -\frac{\phi_* 10^{(\alpha_* + 1)(M_2-M_*)}}{\alpha_* + 1},
\end{equation}
with $\phi_*,~\alpha_*,~M_*$ depending on the redshift \citep[see Table 1 in][]{conselice16}, and $M_2 = 12$.

@@

The time over which these IMRI forms can be estimated as the sum of the cluster formation time, the IMBH formation time, the IMRIs formation time, and the IMRI merger timescale:
\begin{equation}
T = t_{\gc ,f} + t_{\ibh ,f} + t_{\imri, f} + t_{\rm GW}. 
\end{equation}
The physics that regulate IMBH formation is still debated and partly unknown. Depending on the formation scenario, models suggest that the IMBH growth can occur either on short ($\sim 1$ Gyr) or long ($\simeq 5-10$ Gyr) timescales. As discussed recently, {\it fast} IMBHs could outnumber those forming slowly \citep{giersz15,AAG19}, thus in our calculations we assume $t_{\ibh ,f} = 2$ Gyr. 
The formation of an IMRI scales with the mass-segregation timescale in the host cluster, which is expected to be of the order of $\sim 0.1-1$ Gyr, while for $t_\gw$ we calculate, from our models, the value at which the number of mergers is half the total number, i.e. $N_\gw(t_\gw) = 0.5 N_\gw$. This corresponds to $t_\gw = 0.6-1.5$ Gyr.

@@

\section{Conclusions}
\label{end}

In this work we studied the dynamical interaction of an IMBH with two stellar BHs under the assumption
that these three massive object dominate completely the dynamics in the innermost regions of a massive star cluster.
We used $28,000$ direct $N$-body models aimed at exploring the formation and evolution of IMRIs, which in this paper we 
broadly define as tight binaries involving an IMBH with mass $> 100 \Ms$ and a stellar BH with mass $\lesssim 50 \Ms$.
In order to assess the role of environments and dynamics on IMRIs formation, we vary the density and metallicity of the 
host cluster, the IMBH mass, the mass spectrum of stellar BHs, and the orbital properties of the three objects.

>>>>

\titlerunning{Formation of IMRIs in Milky Way globular clusters and the Local Volume}	
\authorrunning{Arca Sedda, Amaro-Seoane, Chen}
\author{Manuel Arca Sedda \inst{1} \and Pau Amaro-Seoane \inst{2,}\inst{3,}\inst{4,}\inst{5} \and Xian Chen \inst{6,}\inst{3} }

\institute{Astronomisches Rechen-Institut. Zentrum f\"{u}r Astronomie der Universit\"{a}t Heidelberg, M\"onchhofstrasse 12-14, Heidelberg, D-69120, DE              
         \and         
			  Universitat Polit{\`e}cnica de Val{\`e}ncia, IGIC, 46022 Valencia, Spain
		\and         
			  Kavli Institute for Astronomy and Astrophysics at Peking University, Beijing 100871, P.R. China
		\and	  
			  Institute of Applied Mathematics, Academy of Mathematics and Systems Science, CAS, Beijing 100190, China
		\and
			  Zentrum f{\"u}r Astronomie und Astrophysik, TU Berlin, Hardenbergstra{\ss}e 36, 10623 Berlin, Germany
		\and
			Astronomy Department, School of Physics, Peking University, Beijing 100871, P.R. China	             
\\		\email{m.arcasedda@gmail.com}			
\\    	\email{amaro@upv.es}
\\		\email{xian.chen@pku.edu.cn }
}

@@

Intermediate mass black holes (IMBH), with masses in the range $10^2-10^5\Ms$, might represent the missing link between stellar and supermassive mass black holes. Dense stellar systems, such as globular clusters (GCs), are thought to be ideal factories for IMBH production via formation and collapse of a very massive stars through stellar collisions \citep{zwart02, freitag06b,freitag06c, giersz15, mapelli16},
or via multiple interactions and mergers between stars and stellar-mass BHs \citep{giersz15}. 
{\bf Aside from the scenario above, several mechanisms have been proposed for IMBH formation. In fact, IMBHs with masses in the range 
$10^2-10^4\Ms$ can form out of the direct collapse of massive stars extremely poor in metals \citep{madau01,bromm02,ohkubo09,spera17} or of gaseous clouds in the early Universe \citep{latif13}, or can form in high redshift, metal poor, galactic halos \citep{bellovary11}. A further possibility is that IMBHs form in the centre of satellite galaxies, later accreted in their host galaxy halo \citep{bellovary10}, or can form in galactic nuclei, either in AGN accretion discs \citep{mckernan12} or in galactic discs circumnuclear regions \citep{taniguchi00}.}

Unfortunately, a striking observational evidence for the presence of IMBHs inhabiting GCs is still missing due to the {\bf impossibility to resolve the region in which they dominate dynamics} \citep[for recent reviews see][]{mezcua17,greene19}. Indeed, models suggest that several processes can mimic an IMBH, like anisotropies inthe GC kinematic  properties \citep{zocchi}, or the presence of a dense cluster of stellar mass BHs -- namely a BH subsystem -- that dominate the inner cluster centre \citep{AAG18b,AS16,vandermarel10}. Nevertheless, a few observational IMBHs candidates have been found for Galactic GCs

@@

properties \citep{Baumgardt17,AAG18a,AAG19}. Due to this, finding an unique way to unravel the presence of IMBHs in GCs represents one of the most interesting challenges in modern astronomy. Beside this, the presence of an IMBH sitting in the centre of a dense cluster represents a scenario particularly appealing from the perspectives of gravitational waves (GW) astronomy. Indeed, a compact remnants orbiting the IMBH can enter a regime where GWs emission dominates, leading to the formation of an intermediate mass-ratio inspiral \citep[IMRI,][]{konstantinidis13,leigh14,haster16,macleod16}, a class of sources possible audible with the next generation of GW observatories like LISA \citep{Will04,seoane07,amaro12,seoane18}. However, in the highly dense regions that characterize star clusters centres, the formation of IMRIs is not a smooth process. Indeed, due to the continuous interactions with stars, an IMRI ``progenitor'', namely a tight IMBH-BH binary, might be subjected to strong perturbation induced, for instance by a passing-by BH. The three-body interaction involving the IMBH and the two BHs can lead to a variety of end states, including the formation of an IMRI, a stellar BH binary, the ejection of one BH, or even both, or the development of a head-on collision. At some extent, this scenario is similar to what is expected to happen in galactic nuclei, where supermassive BHs (SMBHs) can capture a compact object to form an extreme mass ratio inspiral. However, in the case of IMRIs the picture is complicated by the fact that this chaotic process can transfer to the IMBH an amount of energy sufficient to displace it sensibly from the cluster centre. Differently from galactic nuclei, where the SMBH remains well seated in the galactic potential well, the IMBH motion makes hard the use of any analytical approach to solve IMRIs dynamics. Quantifying the branching ratio for IMRIs formation constitute a fundamental step to assess the probability to observe these GW sources with the next generation of space-based detectors like LISA\footnote{\url{https://www.elisascience.org/}} \citep{Will04,seoane07,amaro12}, TianQin \citep{tianqin16} or Taiji \citep{taiji17}. In this paper, we model the potential formation of an IMRI from the interactions between an IMBH and two stellar mass BHs. This type of  configuration can temporarily form in the centre of clusters with IMBH mass $\sim 100-1000\Ms$ \citep{konstantinidis13,macleod16} and could last up to $10^7$ yr \citep{macleod16} although this depends strongly on the cluster initial properties. In fact, it is difficult to predict the actual amount of BHs that, at any time, interact with the IMBH, primarily because the majority of the studies conducted in the literature assume that the IMBH is already at the centre of the host cluster when stellar BHs form and sink toward the cluster centre \citep{leigh14,haster16}. Models in which the IMBH is a byproduct of stellar evolution and collisions, indeed, seem to point out that after the IMBH grows only a handful of stellar BHs are still present in the cluster $<3-5$ \citep{AAG19}. To reach the aim, we use $N$-body simulations that take into account in particles' equations of motion both the star cluster gravitational potential and post-Newtonian corrections at 1, 2, and 2.5 order. Varying the IMBH and BHs masses, their orbital configuration, and the host cluster structural properties, we build-up three sets consisting of 2000 models each, which allow us to uncover three possible scenarios for IMRIs formation.  

@@

{\bf
We simulate the evolution of the innermost region of the cluster, modelling the IMBH and two stellar mass BHs as live particles and
the remaining cluster as an external static potential. In order to explore the parameter space, we create 7 different models, each one consisting of 4000 simulations gathered in four different sub-classes, depending on the IMBH mass, for a total of \manuel{28,000} simulations. Our model is sketched in Figure \ref{f1}.

\begin{figure}
    \centering
    \includegraphics[width=7cm]{triple}
    \caption{Sketch of the IMBH-BH-BH triple configuration. {\bf We mark the main quantities that characterise the triple -- IMBH mass ($M_\ibh$), stellar BH masses ($M_{\rm bh 1,2}$, orbital semimajor axis and eccentricity of the IMBH-BH1 ($a_1,~e_1$) and IMBH-BH2 ($a_2,~e_2$) systems -- and the host cluster -- gravitational potential ($\Phi_\gc$), total mass ($M_\gc$), typical radius ($r_\gc$), and velocity dispersion ($\sigma_\gc$).} }    
\label{f1}
\end{figure}

We vary the central IMBH mass in four classes, namely ${\rm Log}(M_\ibh/\Ms) = 2,~3,~4,~5$. This range covers typical values of putative IMBH masses forming in stellar systems of various sizes, from young and open clusters, to globular clusters, and up to low-mass nuclear clusters and dwarf galaxies.
We note that the lowest value taken for $M_\ibh$ will likely involve mergers with mass ratio $> 0.1$, thus they fall outside the range of IMRIs. Nonetheless, exploring the low end of IMBH mass function will help us in better understanding the perspectives of IMBH-BH mergers from the point of view of both low- \citep[e.g.][]{Will04,seoane18} and high- \citep[e.g.][]{MandelEtAl2008,gair11,abbott17f} frequency GW detectors, as discussed in Section \ref{gra}.

@@

These results have two main implications. \manuel{
Since the IMRI secondary mass reflects the underlying BH mass spectrum, this quantity 
encodes information on the population of BHs lurking in the host cluster. If dynamics had enough time to affect BHs -- e.g. causing the ejection of the most massive via strong scatterings -- or the reservoir of BHs have diminished during the IMBH build-up, the IMRI companion mass distribution will tell us how strongly dynamical processes affected BHs. If, instead, dynamics did not play a major role in shaping the IMBH growth, we would expect that the mass spectrum of IMRI companion would resemble the typical mass spectrum of BHs forming out of a standard stellar initial mass function \citep[e.g.][]{kroupa01}.}

\begin{figure*}
\centering
\includegraphics[width=0.8\textwidth]{massdis}\\
%\includegraphics[width=0.8\textwidth]{massdisD}\\
    \caption{%Top panel: 
    Mass distribution of the IMRI secondary for SET0 to 5. Each color corresponds to a different set to facilitate the comparison between different panels and figures. 
    Dashed lines mark the mass function adopted for stellar BHs.
    }
	\label{fig:f8}
\end{figure*}

@@

\subsection{IMBH survival in dense star clusters}
{\bf
Promptly after a merger event, the anysotropic emission of GWs can impart a recoil kick to the merger product, depending on the mass and spin of the two merging objects \citep{campanelli07,gonzalez07,lousto08,lousto12}. Depending on the kick amplitude, the recoil can kick out the IMBH from the cluster \citep[e.g.][]{bockelmann08,fragione17c}, leading it to wander forever outside the cluster. 
This can significantly affect the retention probability of low-velocity dispersion ($5$ km/s) star clusters, which is limited to $1-5\%$ for BHs with mass $m_\bh \sim 100 \Ms$ that undergo 1-2 consecutive mergers \citep{arca20}.

Assessing the retention probability for IMBHs is crucial to place constraints on their possible presence in star clusters. To explore this aspect, we use our models to create a catalogue of IMRI mergers proceeding on a step-by-step basis as follows: 
\begin{enumerate}
\item we divide the IMBH mass range -- $M_\ibh = 10^2-5\times 10^5\Ms$ -- in 15 values evenly distributed in logarithmic values;
\item for each IMBH mass, we create a sample of 100 stellar BHs with masses evenly drawn in the $M_\bh = 3.5-80 \Ms$ range. \manuel{This conservative choice allows us to explore from a general perspective the possible ejection of the IMBH via GW recoil and cope with our ignorance about the actual mass spectrum of stellar BHs involved in an IMRI.};
\item we associate to the IMBH and BH a spin randomly drawn between 0-1;
\item we use \cite{jimenez17} numerical relativity fitting formulae to calculate the IMRI merger remnant mass $M_r$, spin $S_r$, and recoil velocity $v_r$, following the procedure depicted in \cite{arca20}.
\end{enumerate}
For each $M_\ibh -m_\bh$ pair, we repeat steps 3 and 4 100 times and, for each of them, we compare the recoil velocity with the escape velocity from the cluster centre, which we set to $\sigma = 5$ or $50$ km/s.

@@

\manuel{As in the previous section, for each merger event we calculate the merger remnant mass, spin, and recoil kick using the numerical relativity fitting formulae described in \cite{jimenez17} following \cite{arca20}.} Figure \ref{fig:f11a} shows the value of the remnant IMBH spin as a function of the IMRI mass ratio.

We see that as long as the mass ratio remains below $10^{-2}$ -- which implies an $M_\ibh > 10^4 \Ms$ -- the IMBH spin is not affected by the merger. However, for IMRIs with a mass ratio > 0.01, the remnant IMBH spin can change significantly, with a spin variation $\sim \pm 0.1$ for mass ratios in the range $q = 0.01-0.1$. At mass ratio $q>0.1$, the remnant IMBH spin ($S_r \sim 0.8\pm 0.2$) tends to attain a value close to the average value assumed for the BH spin distribution.
The peculiar dependence between the remnant spin and the IMRI mass ratio has important consequences for the IMBH evolution. One the one hand, if the IMBH formed from the collapse of a gaseous cloud or a very massive star born out of stellar collisions and reach masses $\gtrsim 10^4\Ms$, measuring its spin in a IMRI merger would provide us with information on the processes that regulated the IMBH formation. On the other hand,
if the IMBH formed via multiple mergers in a ''BH-rich'' stellar environment, the remnant spin can provide clues on the spin distribution of the underlying population of stellar BHs.

To further investigate this point, we explore the effects of a series of subsequent mergers on the IMBH final spin. We assume $M_\ibh = 10^3\Ms$, i.e. a mass such to ensure a retention probability of at least $50\%$ in an IMRI merger event involving a BH smaller than $40\Ms$ (see Figures  \ref{fig:survival} and \ref{fig:f11a}). We set the IMBH spin to either $S_\ibh = 0.1-0.5-0.9$. For each spin value, we model ten consecutive mergers with stellar BHs with masses $3-60\Ms$ and spin drawn from a Gaussian distribution with average $S_\bh = 0.5$ and dispersion 0.1. The resulting merger sequence represents an {\it IMBH merger tree}. For each value of the spin we thus create 100 merger trees and calculate the IMBH spin at each successive merger as shown in Figure \ref{fig:f11b}.
We note that all models tend to converge toward the \manuel{same region  of the spin-mass plane, likely close to the spin value adopted for stellar BHs}, thus suggesting that it could be possible to link the IMBH spin with the overall spin distribution of stellar BHs inhabiting the star cluster. In fact, complementing a measure of the IMBH spin with the spin distribution inferred from stellar BH mergers observed as GW sources can help in unravelling the IMBH evolution, whether it was driven by mergers with compact objects or by stellar feeding.
}

@@

\subsection{Gravitational wave signal}
{\bf
Our simulations encompass a wide range of IMRI models, in fact touching the layer of ``ordinary'' BH binaries (mass ratio $q>0.1$) and 
scratching the limit of extreme-mass ratio inspirals ($q < 3 \times10^{-5}$). This setup implies that the GW emission connected with our modelled mergers can populate a wide range of frequencies, from milli- to deca-Hz. Figure \ref{fig:f13} show the characteristic amplitude \citep{kocsis12,ASLiK18,seoane18,robson19} as a function of the frequency for a sample of IMRI mergers with different IMBH mass, assuming they are located at redshift $z=0.1$ ($\sim 460$ Mpc) and an observation time of 4 yr. We compare the modelled signal with the sensitivity curve of several detectors, namely the laser interferometer space antenna \citep[LISA,][]{eLISA13,amaro17lisa}, the advanced laser interferometer antenna \citep[ALIA,][]{bender13}, the decihertz observatory \citep[DO,]{arca19}, the Deciherz Gravitational-Wave Observator \citep[DECIGO,]{Kawamura11}, LIGO \citep{abbott16}, and the Einstein Telescope \citep[ET]{punturo10}. We see that mergers involving IMBHs with masses $< 10^5\Ms$ are promising multiband sources that can be seen in one detector during the inspiral phase and in another during the merger.

@@

The term $\derd M_\gc / \derd M_\ibh$ can be calculated by inverting Equation \ref{mcmibh} and performing the derivative. For the number distribution of the GCs mass in a given galaxy, $\derd n/\derd M_\gc$, we assume a power-law
\begin{equation}
\frac{\derd n}{\derd M_\gc} = k M_\gc^{-s},
\end{equation}
with the slope $s = 2.2$\footnote{{\bf Note that this is compatible with the expected initial mass function of young and old star clusters in galaxies \citep[e.g.][]{gieles09}.}} and the normalization constant 
\begin{equation*}
k = \frac{\delta M_g (2-s)}{(M_{\gc 2}^{2-s} - M_{\gc 1}^{2-s})}.
\end{equation*}
Assuming Galactic values for the galaxy stellar mass, $M_g = 6\times 10^{10}\Ms$, and star clusters mass range, $M_{\gc 1,2} = (5\times 10^3 - 8\times 10^6) \Ms$, we calculate the corresponding IMBH mass range $M_\ibh \simeq (30 - 4.6\times 10^4)\Ms$ according to Equation \ref{mcmibh}, which can also be used to write $M_\gc^{-s} = aM_\ibh^{-bs}$\footnote{The parameters $a,b$ are obtained manipulating Equation \ref{mcmibh}.}. The $\derd n/(\derd M_g\derd z)$ is obtained exploiting the results in \cite{conselice16}, who studied the distribution of galaxies with stellar masses up to $10^{12}\Ms$ up to redshift $z=8$. In particular, we exploit the following parametric expression of galaxies number density 
\begin{equation}
\phi(z) = -\frac{\phi_* 10^{(\alpha_* + 1)(M_2-M_*)}}{\alpha_* + 1},
\end{equation}
with $\phi_*,~\alpha_*,~M_*$ depending on the redshift \citep[see Table 1 in][]{conselice16}, and $M_2 = 12$.

@@

The time over which these IMRI forms can be estimated as the sum of the cluster formation time, the IMBH formation time, the IMRIs formation time, and the IMRI merger timescale:
\begin{equation}
T = t_{\gc ,f} + t_{\ibh ,f} + t_{\imri, f} + t_{\rm GW}. 
\end{equation}
The physics that regulate IMBH formation is still debated and partly unknown. Depending on the formation scenario, models suggest that the IMBH growth can occur either on short ($\lesssim 1$ Gyr) or long ($\simeq 5-10$ Gyr) timescales. As discussed recently, {\bf slowly formed IMBHs could outnumber those forming rapidly by a factor 3-5 \citep{giersz15,AAG19}, thus we assume a weighted formation time defined as $t_{\ibh ,f} = (2/3\times 9) + (1/3\times 0.05) = 2$ Gyr. } The formation of an IMRI scales with the mass-segregation timescale in the host cluster, which is expected to be of the order of $\sim 0.1-1$ Gyr, while for $t_\gw$ we calculate, from our models, the value at which the number of mergers is half the total number, i.e. $N_\gw(t_\gw) = 0.5 N_\gw$. This corresponds to $t_\gw = 0.6-1.5$ Gyr.

@@

\section{Conclusions}
\label{end}
{\bf 
In this work we studied the dynamical interaction of an IMBH with two stellar BHs under the assumption
that these three massive object dominate completely the dynamics in the innermost regions of a massive star cluster.
We used $28,000$ direct $N$-body models aimed at exploring the formation and evolution of IMRIs, which in this paper we 
broadly define as tight binaries involving an IMBH with mass $> 100 \Ms$ and a stellar BH with mass $\lesssim 50 \Ms$.
In order to assess the role of environments and dynamics on IMRIs formation, we vary the density and metallicity of the 
host cluster, the IMBH mass, the mass spectrum of stellar BHs, and the orbital properties of the three objects.
